The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-16 paper/iphas-dr2.tex
Polished Sec 1+2+3

M. J. Barlow$^9$,
P. J. Carter$^{13}$,
R. L. M Corradi$^{6,7}$,
J. J. Drake$^{10}$, \newauthor
J. Eisl\"offel$^{11}$, 
J. Fabregat$^{12}$, 
B. T. G\"ansicke$^{13}$,
N. P. Gentile Fusillo$^{13}$,
S. Greiss$^{13}$, \newauthor
A. S. Hales$^{14}$, 
A. A. Henden$^{15}$,

@@

J. Irwin$^{9}$,

@@

R. Raddi$^{13}$, 

@@

D. Steeghs$^{13}$, 

@@

$^{9}$University College London, Department of Physics \& Astronomy, 

@@

$^{10}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, 

@@

$^{11}$Th\"uringer Landessternwarte, Sternwarte 5, D-07778, Tautenburg, Germany. \\
$^{12}$Observatorio Astron\'omico, Universidad de Valencia,

@@

$^{13}$Department of Physics, University of Warwick, Gibbet Hill Road, Coventry, CV4 7AL, U.K.\\
$^{14}$Joint ALMA Observatory, Alonso de C\'ordova 3107, Vitacura 763-0355, Santiago, Chile. \\
$^{15}$AAVSO, 49 Bay State Road, Cambridge, MA 02138, USA.\\

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights in competitive open time 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007,Wesson2008,Aungwerojwit2012}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011}.

@@

When the initial data release was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to 92 percent of the survey area 
and includes a global photometric re-calibration.
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20 per cent,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only one offset is missing).
        The dark vertical strip near $\ell \sim$125\degr\ 
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in right ascension.}
        \label{fig:footprint}
\end{figure*}

@@

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$.
To cover the Northern Plane with some overlap,
the survey area was divided into 7635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in declination 
and $+$5 arcmin in right ascension,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
Hence, the basic unit of observation hence
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15\,270 fields grouped into 7\,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10~s instead of 30~s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10~s to 20~s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

Fig.~\ref{fig:footprint} shows the footprint
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8 per cent of the survey area --
are predominantly located towards the Galactic anti-centre 
at $\ell > 120\degr$.
Fields at these longitudes are mainly accessed from La Palma 
in the months of November-December,
which is when the La Palma weather and seeing conditions are often poor
forcing many (unsuccessful) repeat observations.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21\arcsec$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied to remove
the contamination from any nearby sources.
This approach works very well 
across the vast majority of the survey area.  Nevertheless,
the Galactic Plane contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
where aperture photometry will be compromised by frequent blending
or poor background subtraction.
In \S\ref{sec:catalogue} we will explain how affected sources are 
flagged. 
%we will explain that objects affected in this way 
%are flagged in the catalogue using the \emph{deblend} warning flag.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74\,195 single-band IPHAS exposures 
in which a total of 1.9~billion \emph{candidate sources} were detected 
at the default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious detections, artefacts and
duplicate detections;
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is the main reason for the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter $r$-band exposure time
that was used in the first year of data-taking.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation of the point spread function (PSF).
It is expected to be zero (circular) across the field 
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
To exclude these, we rejected exposures in which the mean ellipticity
\emph{across the detectors} exceeded $e > 0.3$.
We have also found that, as this threshold is breached,
the photometric measurements delivered by the pipeline become degraded.

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at a resolution better than 1.7~arcsec,
as evaluated by measuring the average PSF Full Width at Half Maximum (FWHM)
across the detectors,
which is a proxy for the combined effect of atmospheric and dome seeing.
This target is currently attained across 86 per cent of the footprint.
To increase the sky area offered by the data release slightly,
we have decided to accept data obtained with seeing up to 2.5~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the PSF FWHM for the approved fields.
In the $r$-band, 90 per cent is better than 1.5~arcsec,
50 per cent is better than 1.1~arcsec,
and 10 per cent is better than 0.8~arcsec
(the WFC pixel scale is 0.33~arcsec/px).
In \S\ref{sec:catalogue} we will explain
that the photometry compiled in the source catalogue
is normally derived from the field with the
best-available seeing for a given object,
and that the PSF FWHM measurement
is included as a column in the catalogue.

@@

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy normally
ensures that every pointing is immediately followed 
by an offset pointing at a displacement
of $+$5~arcmin in Dec and $+$5~arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We rejected field pairs in which more than 2 per cent of the stars 
showed an inconsistent measurement at the level of 0.2~mag,
or more than 26 per cent were inconsistent at the level of 0.1~mag.
These limits were set empirically after inspecting
the images and photometry by eye.

(5) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1~arcsec.
This was particularly effective for revealing
fields with an inaccurate astrometric solution in one of the bands,
which were subsequently corrected.

(6) \emph{Visual examination.}
Images, colour mosaics,
and the associated photometric colour/magnitude diagrams
were inspected by a team of 20 survey members, 
such that each image in the data release 
was looked at by at least three different pairs of eyes.
Images affected by clouds or extreme levels of scattered moonlight
were flagged, investigated,
and excluded from the release 
by placing them on a black-list.

@@

All of the above criteria were satisfied by at least one observing attempt
for 14\,115 out of the 15\,270 fields (92 per cent).
In some cases more than one successful attempt to observe
a field was available due to stricter
quality criteria being applied in the initial years of the survey.
In such cases, only the attempt 
with the best seeing and depth has been selected
for inclusion in the catalogue, in order  
to deliver the most reliable measurement at a single epoch.

We note that some of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The images and detection tables of the discarded data are made
available through our website for this reason (www.iphas.org),
but will be ignored in the remainder of this work.

>>>>

A. A. Henden$^9$,
M. J. Barlow$^{10}$,
P. J. Carter$^{11}$,
R. L. M Corradi$^{6,7}$, \newauthor
J. J. Drake$^{12}$, 
J. Eisl\"offel$^{13}$, 
J. Fabregat$^{14}$, 
B. T. G\"ansicke$^{11}$,
N. P. Gentile Fusillo$^{11}$, \newauthor
S. Greiss$^{11}$, 
A. S. Hales$^{15}$, 

@@

J. Irwin$^{12}$, 

@@

R. Raddi$^{11}$, 

@@

D. Steeghs$^{11}$, 

@@

$^{9}$AAVSO, 49 Bay State Road, Cambridge, MA 02138, USA.\\
$^{10}$University College London, Department of Physics \& Astronomy, 

@@

$^{11}$Department of Physics, University of Warwick, Gibbet Hill Road, Coventry, CV4 7AL, U.K.\\
$^{12}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, 

@@

$^{13}$Th\"uringer Landessternwarte, Sternwarte 5, D-07778, Tautenburg, Germany. \\
$^{14}$Observatorio Astron\'omico, Universidad de Valencia,

@@

$^{15}$Joint ALMA Observatory, Alonso de C\'ordova 3107, Vitacura 763-0355, Santiago, Chile. \\

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights in competitive open time 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7~magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007,Wesson2008,Aungwerojwit2012}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011}.

@@

When the initial data release was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to 92 per cent of the survey area 
and includes a global photometric re-calibration.
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{figures/footprint/footprint_large.jpg}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20 per cent,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that one offset is missing).
        The dark vertical strip near $\ell \simeq$125\degr\ 
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in right ascension.}
        \label{fig:footprint}
\end{figure*}

@@

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10~s instead of 30~s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10~s to 20~s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

Fig.~\ref{fig:footprint} shows the footprint
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8 per cent of the survey area --
are predominantly located towards the Galactic anti-centre 
at $\ell > 120\degr$.
Fields at these longitudes are mainly accessed from La Palma 
in the months of November-December,
which is when the La Palma weather and seeing conditions are often poor,
forcing many (unsuccessful) repeat observations.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21\arcsec$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied to remove
the contamination from any nearby sources.
This approach works very well 
across the vast majority of the survey area.  Nevertheless,
the Galactic Plane contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
where aperture photometry will be compromised by frequent blending
or poor background subtraction.
In \S\ref{sec:catalogue} we will explain how affected sources are 
flagged.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74,195 single-band IPHAS exposures 
in which a total of 1.9~billion \emph{candidate} detections
were made at the default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious objects, artefacts and
duplicate detections;
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a more reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is the main reason for the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
To exclude these, we rejected exposures
in which the \emph{average} ellipticity
across the detectors exceeded $e > 0.3$.
We have also found that, as this threshold is breached,
the photometric measurements delivered by the pipeline become degraded.

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at a resolution better than 1.7~arcsec,
as evaluated by measuring the average PSF Full Width at Half Maximum (FWHM)
across the detectors,
which is a proxy for the image quality
arising from the combined effects of atmospheric and dome seeing.
This target is currently attained across 86 per cent of the footprint.
To increase the sky area offered by the data release slightly,
we have decided to accept data obtained with FWHM up to 2.5~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the PSF FWHM for the approved fields.
In the $r$-band, 90 per cent is better than 1.5~arcsec,
50 per cent is better than 1.1~arcsec,
and 10 per cent is better than 0.8~arcsec
(the WFC pixel scale is 0.33~arcsec/px).
In \S\ref{sec:catalogue} we will explain
that the photometry compiled in the source catalogue
is normally derived from the field with the
best-available seeing for a given object,
and that the FWHM measurement
is available as a column in the catalogue.

@@

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy normally
ensures that every pointing is immediately followed 
by an offset pointing at a displacement
of $+$5~arcmin in Dec and $+$5~arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We automatically rejected field pairs
in which more than 2 per cent of the stars 
showed an inconsistent measurement at the level of 0.2~mag,
or more than 26 per cent were inconsistent at the level of 0.1~mag.
These limits were set empirically after inspecting
the images and photometry by eye.

(5) \emph{Visual examination.}
Images, colour mosaics,
and the associated photometric colour/magnitude diagrams
were inspected by a team of 20 survey members, 
such that each image in the data release 
was looked at by at least three different pairs of eyes.
Images affected by clouds or extreme levels of scattered moonlight
were flagged, investigated,
and excluded from the release 
by placing them on a black-list.

(6) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1~arcsec.
This was effective for revealing
fields with an inaccurate astrometric solution in one of the bands,
which were subsequently corrected.

@@

The above criteria were satisfied by at least one observing attempt
for 14,115 out of the 15,270 planned fields (92 per cent).
In some cases more than one successful attempt to observe
a field was available due to stricter
quality criteria being applied in the initial years of the survey.
In such cases, only the attempt 
with the best seeing and depth has been selected
for inclusion in the catalogue, in order  
to deliver the most reliable measurement at a single epoch.

We note that some of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The images and detection tables of the discarded data are made
available through our website (www.iphas.org),
but will be ignored in the remainder of this work.
